### 2012-03-04 paper_draft.tex
added citations

Although companies like 23andMe are willing to use the data for scientific research,
it is not easy for individual scientists to access the data. 
The generation of biomedical data by private companies raises concerns about privacy \footnote{https://www.23andme.com/about/privacy/}, 
liabilty and consent \footnote{addcitations}. 
Nevertheless, individual customers are willingly sharing their data. Most do so by uploading their data to
their personal website or to open software repositories like \textit{GitHub}. 
This data is scattered and unorganized, making it hard to use this information. While projects like SNPedia try to keep track of all 
the genotyping files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is 
most often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) 
or other methods by which researchers could download the data and not every customer of DTC genetic testing can participate in it.

@@

\subsection*{Survey on Sharing Genetic Information}
In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX participated in our survey on sharing genetic information with the public. 
The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already 
customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% don't plan on 
becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Somers-d).

67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so given the company 
didn't share the data with third parties. 6.6 \% of the participants would not share their data. There is no significant difference in 
readiness to share between chromosomal sexes (Somers-d). Those who are a customers of a DTC company or are planning on becoming one in 
the future are more likely to share their results, compared to those who don't plan on getting themselves genotyped (Somers-d).

There are significant differences in terms of motivation, tested by Tukey's HSD test, between those people who have already been genotyped 
and those who are not planning on getting genotyped. The first group is likely to agree more strongly, on a five-point scale, with the following 
motivations for sharing genotypic information: because they want to help scientists (mean difference = 0.465, SE = 0.128, p = 0.001), 
because they think of personal benefits (mean difference = 0.448, SE = 0.183, p = 0.04) and because they are curious (mean difference = 1.159, SE = 0.193, p < 0.001).

On the other hand, those people who are not planning on getting genotyped are more likely to agree with the following motivations 
for not sharing their data: because they fear discrimination (mean difference = 1.060, SE = 0.195, p < 0.001), because they feel 
it is a breach of their privacy (mean difference = 0.821, SE = 0.225, p = 0.001), because they fear negative consequences for their 
family members (mean difference = 0.733, SE = 0.21, p = 0.002) or because they fear personalized advertising (mean difference = 0.848, SE = 0.208, p < 0.001).

Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with 
the following motivations for sharing than those who would not share their data with their DTC company: because they want to help 
scientists (mean difference = 1.57, SE = 0.199, p < 0.001), because they think of personal benefits (mean difference = 0.951, SE = 0.308, p = 0.006), 
and because they are curious (mean difference = 1.99, SE = 0.321, p < 0.001).

Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the following motivations 
for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they 
fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), 
because they fear consequences for their family members (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).

@@

Despite this benefit in terms of cost GWAS can only detect correlations of SNPs with those traits and don't allow 
inference on the cause for any correlation. Furthermore, GWAS need a large enough sample size, 
which has to be statistically analysed by using sound methods. 
Nevertheless, GWAS are still frequently used and new associations are found. 
(we should list some recent results here -> oxytocin maybe? we could look at the twitter-timeline of @opensnporg, i publish papers i find there).

>>>>

Although companies like 23andMe are willing to use the data for scientific research,
it is not easy for individual scientists to access the data. 
The generation of biomedical data by private companies raises concerns about privacy \footnote{https://www.23andme.com/about/privacy/}, 
liabilty and consent \cite{Caulfield2011}. 
Nevertheless, individual customers are willingly sharing their data. Most do so by uploading their data to
their personal website or to open software repositories like \textit{GitHub}. 
This data is scattered and unorganized, making it hard to use this information. While projects like SNPedia try to keep track of all 
the genotyping files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is 
most often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) 
or other methods by which researchers could download the data and not every customer of DTC genetic testing can participate in it.

@@

\subsection*{Survey on Sharing Genetic Information}
In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX participated in our survey on sharing genetic information with the public. 
The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already 
customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% don't plan on 
becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Cramer's V = 0.077).

67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so given the company 
didn't share the data with third parties. 6.6 \% of the participants would not share their data. Participants self-identified as XX-chromosomal are slightly more likely to answer that DTC companies are allowed to use their results (Cramer's V = 0.221). Those who are a customers of a DTC company or are planning on becoming one in 
the future are more likely to share their results, compared to those who don't plan on getting themselves genotyped (Somers-d = 0.331).

There are substantial differences in terms of motivation, tested by Tukey's HSD test, between those people who have already been genotyped 
and those who are not planning on getting genotyped. The first group is likely to agree more strongly, on a five-point scale, with the following 
motivations for sharing genotypic information: because they want to help scientists (mean difference = 0.465, SE = 0.128), 
because they think of personal benefits (mean difference = 0.448, SE = 0.183) and because they are curious (mean difference = 1.159, SE = 0.193).

On the other hand, those people who are not planning on getting genotyped are more likely to agree with the following motivations 
for not sharing their data: because they fear discrimination (mean difference = 1.060, SE = 0.195), because they feel 
it is a breach of their privacy (mean difference = 0.821, SE = 0.225), because they fear negative consequences for their 
family members (mean difference = 0.733, SE = 0.21) or because they fear personalized advertising (mean difference = 0.848, SE = 0.208).

Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with 
the following motivations for sharing than those who would not share their data with their DTC company: because they want to help 
scientists (mean difference = 1.57, SE = 0.199), because they think of personal benefits (mean difference = 0.951, SE = 0.308), 
and because they are curious (mean difference = 1.99, SE = 0.321). In the case of curiosity as a motive, there is also a substantial difference between those who would share their data with their DTC company under the condition that it did not share the information and those who would not (mean difference = 1.116 SE = 0.344) as well as those who would share under any circumstances (mean difference = -0.874 SE = 0.182).

Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the following motivations 
for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they 
fear discrimination (mean difference = 1.52, SE = 0.322), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324), 
because they fear consequences for their family members (mean difference = 1.146, SE = 0.32) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357). In the cases of fear of discrimination and fear of a breach of privacy, substantial differences between all three categories exist: Those who would share their data with their DTC company as long as it did not share the information agree less strongly than those who would not share the data with both fear of discrimination as a motive for not sharing (mean difference = -0.615, SE = 0.345) as well as fear of a breach of privacy (mean difference = -0.668, SE = 0.346). Those who would share their data under any circumstances are even less likely to agree with these motives than those who would share only if their DTC company did not share the information (fear of discrimination: mean value = -0.906, SE = 0.182; breach of privacy: mean difference = -1.203, SE = 0.183).

@@

Despite this benefit in terms of cost GWAS can only detect correlations of SNPs with those traits and don't allow 
inference on the cause for any correlation. Furthermore, GWAS need a large enough sample size, 
which has to be statistically analysed by using sound methods. 
Nevertheless, GWAS are still frequently used and new associations are found \cite{10.1371.journal.pone.0031470,10.1371.journal.pone.0030309,10.1371.journal.pone.0029848}.
